---
title: "Clustering Airbnb listings in Naples"
output: pdf_document
---

# Introduction

This section outlines the process of cleaning and clustering Airbnb data. The dataset contains various features like price, room type, amenities, and others. We will prepare this data for further analysis and cluster analysis.

## Loading Required Libraries

We start by installing and loading the necessary libraries.

```{r libraries, results='hide'}
# List of CRAN packages
cran_packages <- c(
  "tidyverse", 
  "stringr", 
  "cluster", 
  "fpc", 
  "clustMixType", 
  "geojsonio", 
  "leaflet", 
  "ggplot2", 
  "FactoMineR", 
  "kmed",
  "mgsub"
)

# Install CRAN packages
new_packages <- cran_packages[!(cran_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

library(tidyverse, quietly = TRUE)
library(mgsub)
library(stringr)

# URL of the compressed CSV file
url <- "http://data.insideairbnb.com/italy/campania/naples/2023-06-21/data/listings.csv.gz"

# Temporary file to store the downloaded data
temp_file <- tempfile(fileext = ".csv.gz")

# Download the file
download.file(url, temp_file)

# Read the compressed CSV file into a data frame
airbnb_data <- read_csv(temp_file)

# Clean up
unlink(temp_file)
```

# Data Cleaning

We remove rows where 'price' or 'review_scores_rating' is missing.

```{r sss, results='hide'}
airbnb_data <- drop_na(airbnb_data, price, review_scores_rating)
```

The `gsub` function replaces all occurrences of dollar signs ($) and commas (,) in the price column with an empty string. This is needed because these characters are not valid in a numerical context.

`as.numeric(...)`: The resulting string is then converted to a numeric data type.

`mutate(airbnb_data, price = ...)`: The mutate function from the dplyr package replaces the original price column with the newly transformed numeric column.

```{r sss2, results='hide'}
airbnb_data <- mutate(airbnb_data, price = as.numeric(gsub("[$,]", "", price)))
```

We need to ensure that the date columns are in the correct format for further analysis.

```{r sss3, results='hide'}
# Convert date columns to Date type
airbnb_data <- mutate(airbnb_data, last_review = as.Date(last_review, format="%Y-%m-%d"))
```

We transform the room type into a categorical variable and remove any special characters from the listing names.

```{r sss4, results='hide'}
# Convert 'room_type' to factor, remove special characters from 'name'
airbnb_data <- mutate(airbnb_data, room_type = as.factor(room_type),
                      name = gsub("[^[:alnum:][:space:]]", "", name))
```

Here, we calculate the price per guest and filter out listings with extreme values.

```{r sss5, results='hide'}
# Compute price per guest and filter out extreme values
airbnb_data <- mutate(airbnb_data, price_per_guest = price / accommodates) %>% 
  filter(price_per_guest <= 1000)
```

We extract and convert the bathroom details, including whether the bathroom is shared or private.

```{r sss6, results='hide'}
# Extract and convert bathroom details
airbnb_data <- airbnb_data %>% 
  mutate(
    bathrooms = as.numeric(str_extract(bathrooms_text, "\\d+")),
    bathrooms = if_else(
      str_detect(bathrooms_text, "shared"), 
      bathrooms / 2, 
      bathrooms
    ),
    bathrooms = if_else(
      str_detect(bathrooms_text, "private"), 
      bathrooms, 
      bathrooms
    )
  )
```

We convert percentage strings to numeric values for easier analysis.

```{r sss7, results='hide'}
# Define a function to safely convert the percentage string to numeric
convert_to_numeric <- function(x) {
  num <- as.numeric(str_replace(x, "%", ""))
  ifelse(is.na(num), NA, num / 100)
}
```

We convert host acceptance and response rates to a numerical format. 

```{r sss8, results='hide'}
# Use the function to safely convert the columns
airbnb_data <- airbnb_data %>%
  mutate(
    host_acceptance_rate = suppressWarnings(convert_to_numeric(host_acceptance_rate)),
    host_response_rate = suppressWarnings(convert_to_numeric(host_response_rate))
  )
```

We compute the total number of amenities available for each listing and add this information as a new column.

```{r sss9, results='hide'}
# Compute and add the number of amenities as a new column
airbnb_data <- mutate(airbnb_data, num_amenities = str_count(amenities, ",") + 1)
```

We convert certain date-related columns to Date type and calculate how long the host has been active, the time since the first and last review.

```{r sss10, results='hide'}
# Convert 'host_since', 'first_review', and 'last_review' to Date and calculate durations
current_date <- as.Date("2023-06-21")
airbnb_data <- mutate(airbnb_data, host_since = as.Date(host_since),
                      first_review = as.Date(first_review),
                      last_review = as.Date(last_review),
                      host_length = as.integer(current_date - host_since),
                      since_first_review = as.integer(current_date - first_review),
                      since_last_review = as.integer(current_date - last_review))
```

We create a binary variable to indicate whether a host has multiple listings or not.

```{r sss11, results='hide'}
# Add 'multiple_host_listings' based on the condition
airbnb_data <- mutate(airbnb_data, multiple_host_listings = if_else(calculated_host_listings_count > 1, 1, 0))
```

We create new variables for each distinct amenity and host verification. These variables will be binary (Yes/No) and will help in understanding the importance of each feature for the listings.

We'll be using `mgsub` for multiple string replacements and `stringr` for string manipulations.

First, we clean up the amenities and host verifications strings to make them easier to handle.

```{r sss12, results='hide'}
amenities <- as.data.frame(mgsub(airbnb_data$amenities, c("\\{", "\\}", "\\,", "\\'", "\\[", "\\]", "\"", "\\/") , c("", "", "  ", "", "", "", "", "")))
host_verifications <- as.data.frame(mgsub(airbnb_data$host_verifications, c("\\[", "\\]", "\\,", "\\'") , c("", "", "", "")))
```

We initialize lists to store the amenities and host verifications for each listing.

```{r sss13, results='hide'}
output_list_verifications <- list()
output_list_verifications_count <- list()
output_list_amenities <- list()
output_list_amenities_count <- list()
```

We loop through each listing to populate the lists initialized above.

```{r sss14, results='hide'}
for(i in 1:nrow(host_verifications)){
  output_list_verifications[i] <- str_split(host_verifications[i, ], " ")[1]
  output_list_verifications_count[i] <- length(output_list_verifications[[i]])
  output_list_amenities[i] <- str_split(amenities[i, ], "  ")[1]
  output_list_amenities_count[i] <- length(output_list_amenities[[i]])
}
```


Select the most frequent host verifications to use as column names

```{r sss15, results='hide'}
dummy_host_verifications_cols <- output_list_verifications[[which.max(output_list_verifications_count)]]
```

Create an empty data frame with the same number of rows as airbnb_data and columns as dummy_host_verifications_cols

```{r sss16, results='hide'}
dummy_host_verifications_df <- matrix(nrow = nrow(airbnb_data), ncol = length(dummy_host_verifications_cols)) %>% as.data.frame()
colnames(dummy_host_verifications_df) <- dummy_host_verifications_cols
```

Select the most frequent amenities to use as column names

```{r sss17, results='hide'}
dummy_amenities_cols <- output_list_amenities[[which.max(output_list_amenities_count)]]
```

Create an empty data frame with the same number of rows as airbnb_data and columns as 

```{r sss18, results='hide'}
dummy_amenities_cols
dummy_amenities_df <- matrix(nrow = nrow(airbnb_data), ncol = length(dummy_amenities_cols)) %>% as.data.frame()
colnames(dummy_amenities_df) <- dummy_amenities_cols
```

Combine the original airbnb_data with the dummy data frames

```{r sss19, results='hide'}
airbnb_data <- cbind(airbnb_data, dummy_host_verifications_df, dummy_amenities_df)
```

Populate the host verifications dummy columns

```{r sss20, results='hide'}
for(l in 1:length(output_list_verifications)){
  for(j in 82:84){
    if(colnames(airbnb_data)[j] %in% output_list_verifications[[l]]){
      airbnb_data[l, j] <- "Yes"
    } else{
      airbnb_data[l, j] <- "No"
    }
  }
}
```

Populate the amenities dummy columns

```{r sss21, results='hide'}
for(l in 1:length(output_list_amenities)){
  for(j in 85:ncol(airbnb_data)){
    if(colnames(airbnb_data)[j] %in% output_list_amenities[[l]]){
      airbnb_data[l, j] <- "Yes"
    } else{
      airbnb_data[l, j] <- "No"
    }
  }
}
```

Fill NA values in the new dummy columns with "No"

```{r sss22, results='hide'}
for(j in 82:ncol(airbnb_data)){
  for(i in 1:nrow(airbnb_data)){
    if(is.na(airbnb_data[i, j]) == TRUE){
      airbnb_data[i, j] <- "No"
    } 
  }
  # Convert the new dummy columns to factors
  airbnb_data[, j] <- factor(airbnb_data[, j], levels = c("Yes", "No"))
}
```

Define lists of columns to be converted to factors and numerics

```{r sss23, results='hide'}
factor_cols <- c("instant_bookable", "host_has_profile_pic", "host_identity_verified",
                 "host_is_superhost", "room_type", "property_type", "host_response_time",
                 "host_listings_count", "host_total_listings_count", "neighbourhood_cleansed",
                 "has_availability")

numeric_cols <- c("accommodates", "host_listings_count", "host_total_listings_count", "beds",
                  "minimum_nights", "maximum_nights", "minimum_minimum_nights", "maximum_minimum_nights",
                  "minimum_maximum_nights", "maximum_maximum_nights", "host_length", "since_first_review",
                  "since_last_review", "number_of_reviews", "availability_365", "availability_90",
                  "availability_60", "availability_30", "calculated_host_listings_count")

# Apply the transformations
airbnb_data <- airbnb_data %>% 
  mutate(
    across(all_of(factor_cols), as.factor),
    across(all_of(numeric_cols), as.numeric)
  )
```

Remove unnecessary columns

```{r sss24, results='hide'}
airbnb_data <- airbnb_data |>
  select(-c(id, host_verifications, neighbourhood_group_cleansed, first_review, amenities, number_of_reviews_ltm, number_of_reviews_l30d, last_review, calculated_host_listings_count_entire_homes, calculated_host_listings_count_private_rooms, calculated_host_listings_count_shared_rooms, listing_url, host_picture_url, calendar_updated, calendar_last_scraped, description, neighborhood_overview, neighbourhood, host_neighbourhood, neighbourhood_group_cleansed, picture_url, host_id, host_url, host_thumbnail_url, host_name, host_since, host_location, host_about, scrape_id, last_scraped, license, price, source, name, bathrooms_text))

# Remove rows with missing values
airbnb_data <- airbnb_data[complete.cases(airbnb_data),]
```

```{r ffp}
# Identify numeric and factor variables
numeric_vars <- names(airbnb_data)[sapply(airbnb_data, class) == 'numeric']
factor_vars <- names(airbnb_data)[sapply(airbnb_data, class) == 'factor']

# Create new data frames containing only numeric and factor variables
df_numeric <- airbnb_data[, numeric_vars]
df_factor <- airbnb_data[, factor_vars]

# Remove specific numeric columns that are not needed
df_numeric <- df_numeric |> 
  select(-c("minimum_minimum_nights","maximum_minimum_nights", 
            "minimum_maximum_nights", "maximum_maximum_nights",
            "minimum_nights_avg_ntm","maximum_nights_avg_ntm" ))

# Combine the numeric and factor data frames to create the final airbnb_data (4238 x 135)
airbnb_data <- cbind(df_numeric, df_factor)
```

Calculate average nightly price per guest per neighborhood

```{r leaf}
library(geojsonio)
library(leaflet)

nb_geo <- geojson_read('http://data.insideairbnb.com/italy/campania/naples/2023-06-21/visualisations/neighbourhoods.geojson', what = 'sp')

nb_geo@data$neighbourhood = as.factor(nb_geo@data$neighbourhood)

night_neighbourhood <- airbnb_data %>% group_by(neighbourhood_cleansed) %>% summarize(avg_night_price = mean(price_per_guest)) %>% arrange(desc(avg_night_price))

ggplot(night_neighbourhood[1:10, ], aes(x = reorder(neighbourhood_cleansed, avg_night_price),  y = avg_night_price, 
                                        fill = avg_night_price)) + 
  geom_bar(stat = "identity") + ggtitle("Top 10 most expensive neighbourhoods") + 
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(hjust = 0.5, size = 12),
        axis.text = element_text(size = 12)) + scale_y_continuous(labels = function(x) paste0("$", x)) +
  xlab("") + ylab("Airbnb Price per Night") + coord_flip() + theme(legend.position = "none") + scale_fill_gradient(low = "yellow", high = "red")

colnames(night_neighbourhood)[1] <- "neighbourhood"
nb_geo@data <- left_join(nb_geo@data, night_neighbourhood[, 1:2]) %>% as.data.frame()

for(i in 1:nrow(nb_geo@data)){
  if(is.na(nb_geo@data[i, "avg_night_price"]) == TRUE){
    nb_geo@data[i, 3] <- mean(nb_geo@data[complete.cases(nb_geo@data), 3])
  } else if(is.na(nb_geo@data[i, "avg_night_price"]) == FALSE){
  }
}

pal <- colorNumeric(
  palette = "YlGnBu",
  domain = nb_geo@data$avg_night_price
)

price_per_neighbourhood <- leaflet(nb_geo) %>%
  addTiles() %>% setView(lng = 14.305573, lat = 40.853294, zoom = 10.5) %>%
  addPolygons(stroke = TRUE, fillColor = ~ pal(avg_night_price), fillOpacity = 0.8,
              highlight = highlightOptions(weight = 2,
                                           color = ~ pal(avg_night_price), 
                                           fillOpacity = 1,
                                           bringToFront = TRUE),
              label = ~neighbourhood,
              smoothFactor = 0.2,
              popup = ~ paste(paste(neighbourhood,":"), "<br/>","<b/>", paste("Avg Night Price: ", "$", round(avg_night_price)))) %>%
  addLegend("bottomright", pal = pal, values = ~avg_night_price, opacity = 1.0, 
            title = "Average Airbnb Night Price",
            labFormat = labelFormat(prefix = "$"), na.label="")
price_per_neighbourhood
```

# Clustering

Calculate Gower distance and apply PAM (2 to 10 clusters, returns cluster with best ASW criterion value)

```{r gowerpam, results='hide'}

library(cluster)
library(fpc)
gower_dist <- daisy(airbnb_data, metric = "gower")

pam_fit <- pamk(gower_dist, krange = 2:10, criterion = "asw")
```

Apply K-Prototypes (2 to 6 clusters, 20 reps)

```{r kproto, results='hide'}
library(clustMixType)

kpres2 <- kproto(cbind(df_numeric, df_factor), 2, nstart = 1,verbose = FALSE)
kpres3 <- kproto(cbind(df_numeric, df_factor), 3, nstart = 1,verbose = FALSE)
kpres4 <- kproto(cbind(df_numeric, df_factor), 4, nstart = 1,verbose = FALSE)
kpres5 <- kproto(cbind(df_numeric, df_factor), 5, nstart = 1,verbose = FALSE)
kpres6 <- kproto(cbind(df_numeric, df_factor), 6, nstart = 1,verbose = FALSE)
```

```{r clustersim, results='hide'}
print(table(pam_fit$pamobject$clustering, kpres2$cluster))

require(mclust)
adjustedRandIndex(pam_fit$pamobject$clustering, kpres2$cluster)
```

Calculate Total Variation Distance and apply PAM (3 clusters, returns cluster with best criterion value)

```{r TVD, results='hide'}
library(kmed)
airbnb_dist <- distmix(airbnb_data, method = "ahmad", idnum = 1:32, idcat = 33:103)

#run the sfkm algorihtm on airbnb_dist
(simplekm <- skm(airbnb_dist, ncluster = 3, seeding = 50))

#calculate silhouette of the K-medoids result of airbnb data set 
silairbnb <- sil(airbnb_dist, simplekm$medoid, simplekm$cluster, title = "Silhouette plot of Airbnb data set")

barplotnum(df_numeric[,1:10], simplekm$cluster, alpha = 0.05)

barplotnum(df_numeric[,11:20], simplekm$cluster, alpha = 0.05)

barplotnum(df_numeric[,21:32], simplekm$cluster, alpha = 0.05)

adjustedRandIndex(simplekm$cluster, kpres3$cluster)

#cluster sizes
print(table(kpres3$cluster))
```

We use the leaflet package to display listings from the two groups using lat/long information. This will give us an idea of geographical distribution of the clusters.

```{r dep}
c1 <- cbind(df_numeric, df_factor) %>% 
  filter(simplekm$cluster == 1)

c2 <- cbind(df_numeric, df_factor) %>% 
  filter(simplekm$cluster == 2) 

c3 <- cbind(df_numeric, df_factor) %>% 
  filter(simplekm$cluster == 3)

library(leaflet)
leaflet() %>% setView(lng = 14.305573, lat = 40.853294, zoom = 10) %>%
  addTiles() %>%
  addPolygons(data = nb_geo, color = "#444444", weight = 2, opacity = 1) %>%
  addCircleMarkers(  lng = c1$longitude,
                     lat = c1$latitude,
                     radius = 2,
                     stroke = FALSE,
                     color = "blue",
                     fillOpacity = 0.5,
                     group = "c1"
  ) %>%
  addCircleMarkers(  lng = c2$longitude,
                     lat = c2$latitude,
                     radius = 3,
                     stroke = FALSE,
                     color = "green",
                     fillOpacity = 0.5,
                     group = "c2"
  )%>%
  addCircleMarkers(  lng = c3$longitude,
                     lat = c3$latitude,
                     radius = 3,
                     stroke = FALSE,
                     color = "red",
                     fillOpacity = 0.5,
                     group = "c3"
  )
```

# Describe the clusters

```{r desc}
library(FactoMineR)

desc_clus <- catdes(cbind(airbnb_data,as.factor(simplekm$cluster)),136)
```

```{r desccl1}
head(desc_clus$quanti$`1`)
head(desc_clus$category$`1`)
```

Mod/Cla = Within cluster
Cla/Mod = Across cluster

For the Heating = "Yes":

55.7% of the listings who have Heating belong to Cluster 1

71.4% of the listings who belong to Cluster 1 have Heating

50.2% of the whole population has Heating

The category "Heating=Yes" is over represented (v-test > 0) among listings in the first (blue) cluster whereas "calculated_host_listings_count" is under represented (v-test < 0).

```{r descc22}
head(desc_clus$quanti$`2`)
head(desc_clus$category$`2`)
```

Homework: Describe the clusters or find alternative - and perhaps more interesting - clustering solutions.
